Staged Abstract Interpreters

### 2019-09-14 paper/80_related.tex
polish

\paragraph{Optimizing Static Analysis Through Specialization} The underlying
idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretative overhead of
traversing the syntax tree by specialization. The residual programs of AC can
be either textual (using quasi-quotations) or closures (using higher-order
functions).  Our approach is more flexible between representations and requires
less engineering efforts to refactor the analyzer (Section \ref{cs_ac}).
\citet{Johnson:2013:OAA:2500365.2500604} adapted closure generation to optimize a
small-step abstract interpreter in state-transition style. The program
is firstly compiled to an ``abstract bytecode'' IR. The IR consists of higher-order
functions, which will be executed later on an abstract abstract machine for
that IR.  \citet{damian1999partial} provided a formal treatment for abstract
compilation and control-flow analysis, as well as proofs to establish the
correctness of a certified specialized analyzer.  \citet{amtoft1999partial}
applied partial evaluation for constraint-based control-flow analysis.

@@

\paragraph{Construction of Abstract Interpreters} Abstract interpretation is a
semantics-based approach to build sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77, Cousot98-5}.
As for building semantic artifacts, the Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach shows that
abstract interpreters can be derived systematically from concrete semantic
artifacts based on operational semantics.
The AAM approach is closely related to control-flow analysis for higher-order languages
\cite{Midtgaard:2012:CAF:2187671.2187672, Shivers:1991:SSC:115865.115884}.
Using monads to construct abstract interpreters was explored by
\citet{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17,
Darais:2015:GTM:2814270.2814308}.
The unstaged abstract interpreter in this paper follows the abstracting
definitional interpreters (ADI) approach \cite{DBLP:journals/pacmpl/DaraisLNH17}.  
Similar to ADI, reconstructing
big-step abstract interpreters with delimited continuations was shown by
\citet{Wei:2018:RAA:3243631.3236800}; \citet{Keidel:2018:CSP:3243631.3236767}
presented a modular, arrow-based abstract interpreter that allows soundness
proof to be constructed compositionally.
\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis, from where we borrow the notation of modular analysis.
However, the staged abstract interpreter does not generate complete summaries for
modules; instead, it specializes the analysis modularly. Similarly, AC
is considered as an example of this kind of modular analyses \cite{DBLP:conf/cc/CousotC02}.

@@

\paragraph{Partial Evaluation and Multi-stage Programming}
Partial evaluation is an automatic technique for program
specialization and was studied comprehensively by
\citet{10.1007/3-540-61580-6_11, DBLP:books/daglib/0072559}.
In this paper, we use mutli-stage programming to specialize programs.
The Lightweight Modular Staging (LMS) framework
\cite{DBLP:conf/gpce/RompfO10} we used in the paper relies on
type-level stage annotations.  Other notable MSP systems using syntactic
quotations exist in the ML family, e.g., MetaML \cite{DBLP:conf/pepm/TahaS97}
and MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}. 
Compared with other approaches using syntactic quotations, LMS provides
guarantees for preserving evaluation orders within a stage.
\citet{Ofenbeck:2017:SGP:3136040.3136060} and
\citet{Amin:2017:CTI:3177123.3158140} discussed staging for generic programming
as well as stage polymorphism in LMS.
%The idea of staging an abstract interpreter we presented here
%is also applicable to other MSP systems.
Multi-stage programming has been widely used to improve the
performance in many other domains, such as optimizing compilers or
domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14,
  DBLP:conf/snapl/RompfBLSJAOSKDK15,
  DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
  DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
  DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17,
  Ofenbeck:2017:SGP:3136040.3136060}, data processing
\cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
  DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.

As a source of inspiration of this paper, the Futamura projections reveal a
close relation between interpreters and compilers, which was
originally proposed by \citet{futamura1971partial}, and later reprinted \cite{Futamura1999}.
\citeauthor{Amin:2017:CTI:3177123.3158140} considered a tower of concrete
interpreters and showed how to collapse them by using MSP -- it would be interesting
to explore this idea for multiple abstract interpreters
\cite{Cousot:2019:AAI:3302515.3290355, Giacobazzi:2015:APA:2676726.2676987}.
\citet{10.1007/11561347_18} discussed a monad with let-insertion and
memoization for code generation, using Gaussian Elimination in MetaOCaml as an
example. In this work, let-insertion is handled by the ANF transformation in
LMS, and the memoizing monad is just a combination of a reader monad and a state monad
operating on staged data.  Similar to the idea in this paper, specializing
monadic interpreters was explored by \citet{danvy1991compiling,
DBLP:conf/dsl/SheardBP99}. However, they are different from the performance
motivation here.

>>>>

\paragraph{Optimizing Static Analysis Through Specialization} The underlying
idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretative overhead of
traversing the syntax tree by specialization. The residual programs of AC can
be either textual (using quasi-quotations) or closures (using higher-order
functions).  Our approach is more flexible between representations and requires
less engineering efforts to refactor the analyzer (Section \ref{cs_ac}).
\citet{Johnson:2013:OAA:2500365.2500604} adapted closure generation to optimize a
small-step abstract interpreter in state-transition style. The program
is first compiled to an ``abstract bytecode'' IR. The IR consists of higher-order
functions, which will be executed later on an abstract abstract machine for
that IR.  \citet{damian1999partial} provided a formal treatment for abstract
compilation and control-flow analysis, as well as proofs to establish the
correctness of a certified specialized analyzer.  \citet{amtoft1999partial}
applied partial evaluation for constraint-based control-flow analysis.

@@

\paragraph{Construction of Abstract Interpreters} Abstract interpretation is a
semantics-based approach of sound static analyses by approximation
\cite{DBLP:conf/popl/CousotC77, Cousot98-5}.
As for building semantic artifacts, the Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach shows that
abstract interpreters can be derived systematically from concrete semantic
artifacts based on operational semantics.
The AAM approach is closely related to control-flow analysis for higher-order languages
\cite{Midtgaard:2012:CAF:2187671.2187672, Shivers:1991:SSC:115865.115884}.
Using monads to construct abstract interpreters was explored by
\citet{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17,
Darais:2015:GTM:2814270.2814308}.
The unstaged abstract interpreter in this paper follows the abstracting
definitional interpreters (ADI) approach \cite{DBLP:journals/pacmpl/DaraisLNH17}.  
Similar to ADI, reconstructing
big-step abstract interpreters with delimited continuations was shown by
\citet{Wei:2018:RAA:3243631.3236800}.
\citet{Keidel:2018:CSP:3243631.3236767}
presented a modular, arrow-based abstract interpreter that allows soundness
proofs to be constructed compositionally.  \citet{DBLP:conf/cc/CousotC02}
proposed an abstract interpretation framework for modular analysis. In Section
\ref{modular}, we borrow the notation of modular analysis.  However, the staged
abstract interpreter does not generate complete summaries for modules; instead,
it specializes the analysis modularly. Similarly, AC is considered as an
example of this kind of modular analyses \cite{DBLP:conf/cc/CousotC02}.

@@

\paragraph{Partial Evaluation and Multi-stage Programming}
Partial evaluation is an automatic technique for program
specialization \cite{10.1007/3-540-61580-6_11, DBLP:books/daglib/0072559}.
In this paper, we use mutli-stage programming to specialize programs.
The Lightweight Modular Staging (LMS) framework
\cite{DBLP:conf/gpce/RompfO10} we used in the paper relies on
type-level stage annotations.  Other notable MSP systems using syntactic
quotations exist in the ML family, e.g., MetaML \cite{DBLP:conf/pepm/TahaS97}
and MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}. 
Compared with other approaches using syntactic quotations, LMS provides
guarantees for preserving evaluation orders within a stage.
\citet{Ofenbeck:2017:SGP:3136040.3136060} and
\citet{Amin:2017:CTI:3177123.3158140} discussed staging for generic programming
as well as stage polymorphism in LMS.
%The idea of staging an abstract interpreter we presented here
%is also applicable to other MSP systems.
Multi-stage programming has been widely used to improve the
performance in many other domains, such as optimizing compilers or
domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14,
  DBLP:conf/snapl/RompfBLSJAOSKDK15,
  DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
  DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
  DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17,
  Ofenbeck:2017:SGP:3136040.3136060}, data processing
\cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
  DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18, Rompf:2015:FPS:2784731.2784760},
etc.

As a source of inspiration of this paper, the Futamura projections reveal a
close relation between interpreters and compilers, which was
originally proposed by \citet{futamura1971partial} (reprinted as \cite{Futamura1999}).
\citet{Amin:2017:CTI:3177123.3158140} considered a tower of concrete
interpreters and showed how to collapse them by using MSP -- it would be interesting
to explore this idea for multiple abstract interpreters
\cite{Cousot:2019:AAI:3302515.3290355, Giacobazzi:2015:APA:2676726.2676987}.
\citet{10.1007/11561347_18} discussed a monad with let-insertion and
memoization for code generation, using Gaussian Elimination in MetaOCaml as an
example. In this work, let-insertion is handled by the ANF transformation in
LMS, and the memoizing monad is just a combination of a reader monad and a state monad
operating on staged data.  Similar to the ideas in this paper, specializing
monadic interpreters was explored by \citet{danvy1991compiling} and
\citet{DBLP:conf/dsl/SheardBP99}. However, these work are different from the performance
motivation here.

### 2019-09-14 paper/paper.tex
conc
subtitle
polish

%% Title information
%\title{Staged Abstract Interpreters}         %% [Short Title] is optional;
\title{Staged Abstract Interpreters}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Fast and Compositional Whole-Program Analysis for Free}                     %% \subtitle is optional
\subtitle{Fast and Modular Whole-Program Analysis via Meta-programming}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

\iffalse
\author
[Guannan Wei, Yuxuan Chen, Tiark Rompf]
{
\vspace{-2ex}
Guannan Wei, Yuxuan Chen, Tiark Rompf\\
Purdue University
\vspace{-0.5ex}
}
\fi

@@

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  We thank the reviewers of PLDI '19 and OOPSLA '19, as well as James Decker,
  Guanhong Tao, Gr\'egory Essertel, and Fei Wang for their feedback on the early
  drafts. We thank David Van Horn and David Darais for their encouragement on
  this work.
  This work was supported in part by NSF awards 1553471, 1564207, 1918483, DOE
  award DE-SC0018050, as well as gifts from Google, Facebook, and VMware.

>>>>

%% Title information
%\title{Staged Abstract Interpreters}         %% [Short Title] is optional;
\title{Staged Abstract Interpreters}         %% [Short Title] is optional;
\subtitle{Fast and Modular Whole-Program Analysis via Meta-programming}                     %% \subtitle is optional
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  We thank the reviewers of PLDI '19 and OOPSLA '19, as well as James Decker,
  Guanhong Tao, Gr\'egory Essertel, Fei Wang, Qianchuan Ye, Yapeng Ye, and Le Yu
  for their feedback on early drafts. We thank David Van Horn and David
  Darais for their encouragement on this work.
  This work was supported in part by NSF awards 1553471, 1564207, 1918483, DOE
  award DE-SC0018050, as well as gifts from Google, Facebook, and VMware.
